---
output:
  pdf_document:
    number_sections: true
    dev: png
    latex_engine: xelatex
header-includes: \setlength{\emergencystretch}{3em}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(out.width = '100%')
knitr::opts_chunk$set(fig.width=12, fig.height=8)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
```

# Load packages
```{r}
suppressPackageStartupMessages({
  library(GEOquery)
  library(wateRmelon)
  library(dplyr)
  library(msqrob2)
  library(ChAMPdata)
})
```

# Intensity values

```{r}
infdata <- readEPIC("data/methylation") 
```


```{r}
dim(betas(infdata))
```

```{r}
head(betas(infdata))
```

```{r}
dim(exprs(infdata))
```

```{r}
head(exprs(infdata))
```


# Annotation

We will be investigating differential DNA methylation between samples from patients with different disease states.

```{r}
# Read table and clean up
meth_annotation <- getGEO(filename="GSE179325_series_matrix.txt.gz")
meth_annotation <- pData(meth_annotation)
meth_annotation <- meth_annotation %>% 
  dplyr::rename(Disease_state = `disease state:ch1`,
                Tissue_type = `tissue/cell type:ch1`,
                Organism = `organism_ch1`,
                Age = `age:ch1`,
                Gender = `gender:ch1`)
meth_files <- list.files("data/methylation/")
meth_files <- meth_files[grep("Grn.idat",meth_files)]
meth_samples <- unlist(strsplit(meth_files,"_") %>% vapply('[','',1))
meth_annotation <- dplyr::filter(meth_annotation, geo_accession %in% meth_samples)
```

```{r}
# Annotation data
print(meth_annotation[c("geo_accession", "Disease_state", "Age", "Gender", 
                        "Tissue_type", "Organism")])
```

# Preprocessing

## Remove NA values

```{r}
infdata <- infdata[rowSums(is.na(exprs(infdata)))==0,]
```

## Filtering

```{r}
#Remove probes for which calling p-value is insufficient
infdata.pf <- pfilter(infdata)
#change sample names
sampleNames(infdata) <- paste(meth_annotation["geo_accession"][,1],
                              meth_annotation["Disease_state"][,1],sep="_")
```
17430 sites were removed as beadcount <3 in 5% of samples.
And no sites had 1% of samples with a detection p-value greater than 0.05.

## Degree of methylation between groups

```{r}
boxplot(betas(infdata),las=2)
```

There does not seem to be a big difference between groups.

We perform a t-test to confirm.

```{r}
meth_mean_SEVERE <- rep(0,5)
meth_mean_NEGATIVE <- rep(0,5)
n <- 1
m <- 1
for (i in 1:ncol(infdata)){
  if (meth_annotation["Disease_state"][,1][i] == "SEVERE"){
  meth_mean_SEVERE[n] <- mean(betas(infdata)[,i])
  n <- n + 1
  } 
  else {
  meth_mean_NEGATIVE[m] <- mean(betas(infdata)[,i])
  m <- m + 1
}
}
```

```{r}
t_test_res <- t.test(meth_mean_SEVERE,meth_mean_NEGATIVE)
t_test_res
```

The t-test also shows the same result as the previous box plot.
The p-value is 0.9924, above 0.05. Therefore, we cannot reject null hypothesis, which states that there is no difference between groups in terms of average methylation.

```{r}
dat_boxplot <- data.frame(betas = c(meth_mean_NEGATIVE, meth_mean_SEVERE),
                                    group = meth_annotation$Disease_state)
par(mfrow=c(1,2))
boxplot(betas~group,dat_boxplot,las=2)
boxplot(betas~group,dat_boxplot,las=2,ylim=c(0,1))
```

The box plots also show that there is little difference between average methylation levels between groups. 

## Normalisation & Quality Control

```{r}
# Perform normalization including dye color adjustment
infdata.dasen.pf <- dasen(infdata.pf)
# Make methylumi objects to check density and color bias adjustment
infdataM <- as(infdata.pf, "MethyLumiM")
infdataN <- as(infdata.dasen.pf, "MethyLumiM")
```

```{r}
# Make QC plot
par(mfrow=c(2,2))
plotColorBias1D(infdataM,channel="both",main="before")
plotColorBias1D(infdataN,channel="both",main="after")
density(infdataM,xlab="M-value",main="before")
density(infdataN,xlab="M-value",main="after")
```

# Differential analysis

## MDS plot

```{r}
disease <- as.factor(meth_annotation$Disease_state)
par(mar=c(6,6,6,6))
plotMDS(infdataN, labels=paste(meth_annotation$Gender,meth_annotation$Age,sep='-'),
        col=as.double(disease))
par(xpd=T)
legend(par("usr")[2]*-0.5,par("usr")[4]*1,sort(unique(disease)),
       pch=c(16),col=as.double(sort(unique(disease))))
```

There seems to be quite a big gender effect in the samples.

Surprisingly, there does not seem to be a big age effect. However, only a few samples are used in the analysis.
And in the paper of the dataset, the authors report that they find a large age effect, larger than that of the gender effect.

Therefore, we add both gender and age as the confounding factor.

## Design matrix

```{r}
disease <- as.factor(meth_annotation$Disease_state)
age <- as.numeric(meth_annotation$Age)
gender <- as.factor(meth_annotation$Gender)
```

```{r}
design <- model.matrix(~disease+gender+age)
design
```

We do not consider interaction terms so the analysis does not become too complex.

```{r}
cont.matrix <- makeContrast("diseaseSEVERE=0", parameterNames = colnames(design))
cont.matrix
```

## Limma

```{r}
fit <- lmFit(infdataN,design)
fit2 <- contrasts.fit(fit,cont.matrix)
fit2 <- eBayes(fit2)
```

## Results

```{r}
# DE results
LIMMAout <- topTable(fit2,adjust="BH",number=nrow(exprs(infdataN)))
head(LIMMAout)
## Check M-values for top results
exprs(infdataN)[rownames(infdataN)%in%rownames(head(LIMMAout)),]
```


# Plots

## Volcano plot

```{r}
volcano<- ggplot(LIMMAout,aes(x=logFC,y=-log10(P.Value),color=adj.P.Val < 0.05)) + 
          geom_point() + scale_color_manual(values=c("black","red"))
print(volcano)
```

## MA plot


```{r}
MA<- ggplot(LIMMAout,aes(x=AveExpr,y=logFC,color=adj.P.Val < 0.05)) + 
            geom_point() + scale_color_manual(values=c("black","red"))
print(MA)
```

# Functional annotation of results

## Load annotation

```{r}
data("probe.features.epic")
annotation_MA <- probe.features
print(head(annotation_MA))
```

## Check annotation

```{r}
# sort probes
annotation_MA <- annotation_MA[sort(rownames(annotation_MA),index.return=T)$ix,]
# Check if all probes are present in both sets
dim(LIMMAout)
dim(annotation_MA)
sum(LIMMAout$Probe_ID%in%rownames(annotation_MA))
sum(rownames(annotation_MA)%in%LIMMAout$Probe_ID)
```

The dimensions for the annotation is not the same as our output, so we remove probes not found in our results.

```{r}
annotation_MA <- annotation_MA[rownames(annotation_MA)%in%LIMMAout$Probe_ID,]
dim(annotation_MA)
```

## Annotate results

```{r}
# Sort LIMMA output alphabetically on probe name
LIMMAout_sorted <- LIMMAout[sort(LIMMAout$Probe_ID,index.return=T)$ix,]
# Add gene names to LIMMA output
LIMMAout_sorted$Gene <- annotation_MA$gene
LIMMAout_sorted$Feature <- annotation_MA$feature
LIMMAout_sorted$Chrom <- annotation_MA$CHR
LIMMAout_sorted$Pos <- annotation_MA$MAPINFO
LIMMAout_sorted$Chrom <- as.character(LIMMAout_sorted$Chrom)
LIMMAout_sorted$Gene <- as.character(LIMMAout_sorted$Gene)
LIMMAout_sorted$Feature <- as.character(LIMMAout_sorted$Feature)
```

# Quantification of absolute methylation differences

```{r}
dim(LIMMAout_sorted)
dim(betas(infdata.dasen.pf))
```

The dimensions are the same, so we move on.

## Add gene names

```{r}
LIMMAout_sorted$SEVERE_meth <- rowMeans(betas(infdata.dasen.pf)[rownames(infdata.dasen.pf)
    %in%LIMMAout_sorted$Probe_ID, meth_annotation["Disease_state"][,1]=="SEVERE"])
LIMMAout_sorted$NEGATIVE_meth <- rowMeans(betas(infdata.dasen.pf)[rownames(infdata.dasen.pf)
    %in%LIMMAout_sorted$Probe_ID, meth_annotation["Disease_state"][,1]=="NEGATIVE"])
LIMMAout_sorted$Abs_diff_meth <- abs(rowMeans(betas(infdata.dasen.pf)[rownames(infdata.dasen.pf)
    %in%LIMMAout_sorted$Probe_ID, meth_annotation["Disease_state"][,1]=="SEVERE"]) -
    rowMeans(betas(infdata.dasen.pf)[rownames(infdata.dasen.pf)
    %in%LIMMAout_sorted$Probe_ID, meth_annotation["Disease_state"][,1]=="NEGATIVE"]))
```

We also sort the results again on p-values.

```{r}
LIMMAout_annot <- LIMMAout_sorted[sort(LIMMAout_sorted$P.Value,index.return=T)$ix, 
                                  c(1,12,13,10,11,4,7,8,5,14,15,16)]
```

# Interpretation

## Genic regions

```{r}
sum(LIMMAout_annot$adj.P.Val<0.05)
sum(LIMMAout_annot$adj.P.Val[LIMMAout_annot$Gene!=""]<0.05)
head(LIMMAout_annot[c(4,5,6,8,10,11,12)])
```

```{r}
LIMMAout_annot_gene <- LIMMAout_annot[LIMMAout_annot$Gene!="",]
checkMeth <- LIMMAout_annot_gene %>% filter(adj.P.Val < 0.05 & 
                                              abs(LIMMAout_annot_gene$logFC)>2)
checkMeth[c(4,5,6,8,10,11,12)]
```

We only find 1 differentially methylated site, at gene `FAM38A`.

## Promoter regions

```{r}
LIMMAout_annot_prom <- LIMMAout_annot_gene[grepl("TSS",LIMMAout_annot_gene$Feature) | 
                                        (LIMMAout_annot_gene$Feature=="1stExon"),]
head(LIMMAout_annot_prom)
head(LIMMAout_annot_prom %>% filter(LIMMAout_annot_prom$adj.P.Val<0.1))
```

We have no differentially methylated promoter regions.

